---
title: "Car Selling Price"
output: 
  html_document:
    theme: cerulean
    code_folding: hide
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(ggplot2)
library(tidyverse)
library(dplyr)
library(knitr)
library(MASS)

car <- read.csv("carselling.csv")
```

# How does vehicle price change as mileage increases?

I collected price and mileage data for vehicles similar to mine from a car sales website "https://www.ksl.com/auto" and saved it as a CSV file.

## Hypotheses
$$
H_0: \beta_1 = 0 \quad \text{(Mileage has no relationship with price)}
$$

$$
H_a: \beta_1 \neq 0 \quad \text{(Mileage is associated with price)}
$$


## Graph

```{r}
ggplot(car, aes(x=mileage, y=cost))+
  geom_point(alpha=0.6)+
  geom_smooth(method="lm", color="darkblue",se = FALSE, linewidth = 1)+
  geom_point(aes(x = 80000, y = 9000), color = "orange", size = 5, shape = 18)+
  annotate("text", x=80000, y=9000, label="Bought")+
  theme_minimal()
```


## Y-transformation 

$$ \prime{Y}= log(\text{cost})$$
The range of lambda is around 0 to -0.5. There is 0 and -1 so, I decided to use 0 value to maximize the likelihood. 

```{r boxcox, echo=TRUE}
c.lm <- lm(cost~mileage, data=car)
#boxCox(c.lm)
```


```{r}
c_log <- lm(log(cost)~mileage, data =car)
b0 <- coef(c_log)[1]
b1 <- coef(c_log)[2]

buy_miles  <- 80000
buy_price  <- 9000
sell_miles <- 180000
sell_price_hat <- exp(predict(c_log, newdata = data.frame(mileage = sell_miles)))

```


```{r}
sell_price_hat <- exp(predict(c_log, newdata = data.frame(mileage = sell_miles)))

ggplot(car, aes(x = mileage, y = cost)) +
  geom_point(alpha = 0.6) +

  stat_function(
    fun = function(x) exp(b0 + b1 * x),
    linewidth = 1
  ) +

  geom_point(aes(x = buy_miles, y = buy_price),
             shape = 8, size = 5) +
  geom_point(aes(x = sell_miles, y = sell_price_hat),
             shape = 8, size = 5) +

  geom_segment(
    aes(x = buy_miles, y = buy_price,
        xend = sell_miles, yend = sell_price_hat),
    linetype = "dashed", linewidth = 1
  ) +

  annotate("label", x = buy_miles, y = buy_price,
           label = "Bought here") +
  annotate("label", x = sell_miles, y = sell_price_hat,
           label = "Plan to sell here") +

  labs(
    title = "Vehicle Price vs Mileage",
    subtitle = "Slope between  and ★ represents average purchase cost per mile",
    x = "Mileage (miles)",
    y = "Price (USD)"
  ) +
  theme_minimal()

```

## Prediction

```{r}
predict(c_log, data.frame(mileage=150000))
exp(8.685233)
predict(c.lm,data.frame(mileage=150000))

```

## Dignostics

```{r, fig.height=3}
par(mfrow=c(1,3))
plot(c.lm,which=1:2)
plot(c.lm$residuals)
```

```{r, fig.height=3}
par(mfrow=c(1,3))
plot(c_log,which=1:2)
plot(c_log$residuals)
```

# Conclusion

Using a log-transformed linear regression model, vehicle price was found to decrease as mileage increases. 
Based on the fitted model, the predicted selling price at 150,000 miles is approximately 5914.92 dollars.

The average purchase cost per mile, considering only the purchase price, is:

$$\
\frac{9000 - \widehat{\text{Sell Price}}}{180000 - 80000}
$$

This indicates that each additional mile driven costs approximately
```{r}
round((9000 - sell_price_hat)/(150000 - 80000), 3)
```

dollars in purchase cost alone.

From a purchase-cost perspective, selling later reduces the average cost per mile, though operating costs were not considered in this analysis.



## prediction intervals 

```{r}
p <- 150000
mypr <- predict(c.lm, data.frame(mileage=p), interval="prediction")
mypr
b <- coef(c_log)
```

```{r}
ggplot(car, aes(x=mileage, y=cost))+
  geom_point()+
  stat_function(fun=function(x) exp(b[1]+b[2]*x))+
  geom_segment(x=p,xend=p, y=mypr[2], yend=mypr[3], color="red", alpha=0.01, lwd=4)+
  theme_classic()
```


